---
title: "1. Joe Model Overview"
output: rmarkdown::html_vignette
date: '`r Sys.Date()`'
vignette: >
  %\VignetteIndexEntry{1. Joe Model Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(kableExtra)
```

# Overview

This tutorial provides an introduction to the **Joe Model**, a core component of the CEMPRA (Cumulative Effects Model for Prioritizing Recovery Actions) framework. By the end of this tutorial, you will understand how to:

- Import stressor-response and stressor-magnitude data
- Calculate system capacity for individual stressors
- Run the full Joe Model to assess cumulative effects across multiple stressors

## What is the Joe Model?

The Joe Model is a **cumulative effects assessment tool** that quantifies how multiple environmental stressors combine to affect the capacity of a system to support a target species or ecosystem function.

The model works by:

1. **Defining stressor-response relationships**: How does each environmental stressor (e.g., temperature, sediment, flow alteration) affect system capacity?
2. **Measuring stressor magnitudes**: What are the current levels of each stressor at locations of interest?
3. **Combining effects**: How do multiple stressors interact to produce a cumulative effect on the system?

The output is a **system capacity score** (0-100%) representing the relative ability of each location to support the target species, given current stressor conditions.

## Prerequisites

Users following this tutorial should have:

- R and RStudio installed
- The `devtools` package installed
- Basic familiarity with R data structures (data frames, lists)

# Installation

The CEMPRA R package is hosted on GitHub and can be installed using `devtools`.

```{r install, message=FALSE, warning=FALSE}
library(devtools)
# Uncomment and run the following line to install the CEMPRA package
# install_github("essatech/CEMPRA")
library(CEMPRA)
```

# Understanding the Input Data

The Joe Model requires two Excel workbooks as input:

| Workbook | Purpose |
|----------|---------|
| **Stressor-Response Workbook** | Defines how each stressor affects system capacity (the dose-response curves) |
| **Stressor-Magnitude Workbook** | Contains measured or estimated stressor levels at each location |

These two workbooks work together: the stressor-response workbook tells us *how bad* a given stressor level is, while the stressor-magnitude workbook tells us *what the stressor levels actually are* at each location.

# Stressor-Response Workbook

The stressor-response workbook defines the relationship between environmental conditions and system capacity. Think of it as answering the question: *"If temperature is 18°C, what percentage of optimal habitat capacity remains?"*

## Workbook Structure

The stressor-response Excel workbook must follow a standardized format with:

1. A **Main** worksheet listing all stressors and their properties
2. **Individual worksheets** for each stressor containing the dose-response curve data

## The Main Worksheet

The `Main` worksheet serves as a master index of all stressors in your analysis. Here's an example of what it looks like:

```{r, format_demo1, echo = FALSE, message=FALSE, warning=FALSE}
  filename_sr <- system.file("extdata", "stressor_response_fixed_ARTR.xlsx", package = "CEMPRA")
  srw <- StressorResponseWorkbook(filename = filename_sr)
  df <- data.frame(head(srw$main_sheet, 5))
  print(df)
```

Each row represents one stressor, with the following columns:

| Column | Description |
|--------|-------------|
| **Stressors** | Unique name for the stressor (must match worksheet name exactly) |
| **Stressor_cat** | Category grouping (used in population model for linking to vital rates) |
| **Interaction** | How this stressor combines with others: `NA`/blank (multiplicative), `Additive`, or `Minimum` |
| **Linked** | For minimum interactions, which stressor this one is linked to |
| **Stress_Scale** | Scale of the x-axis: `linear` or `logarithmic` |
| **Function** | Curve type: `continuous` (smooth interpolation) or `step` (discrete thresholds) |
| **Life_stages** | Which life stage this stressor affects (e.g., `fry_parr`, `adult`) |
| **Parameters** | What the stressor affects: blank/`NA` for system capacity, or `survival`/`capacity` for vital rates |

## Stressor-Response Curve Worksheets

After the Main worksheet, each stressor must have its own worksheet containing the dose-response data. **The worksheet name must exactly match the stressor name in the Main sheet.**

Each dose-response curve defines how system capacity changes across a range of stressor values:

```{r, format_demo2, echo = FALSE, message=FALSE, warning=FALSE}
  dft <- data.frame(Temperature = c(12, 14, 16, 18, 20, 22, 24, 26, 28),
             msc = c(60, 80, 100, 95, 30, 15, 0, 0, 0),
             SD = c(15, 10, 10, 10, 5, 3, 0, 0, 0),
             low.limit = c(10, 30, 40, 40, 0, 0, 0, 0, 0),
             up.limit = c(80, 95, 100, 100, 50, 30, 0, 0, 0)
   )
  dft %>%
  kableExtra::kbl(caption = "Example dose-response curve for temperature and adult system capacity",
                  col.names = c("Temperature (°C)",
                           "Mean System Capacity (%)",
                           "SD",
                           "Lower Limit",
                           "Upper Limit")) %>%
  kableExtra::kable_classic(full_width = F, html_font = "Cambria")
```

The columns are:

| Column | Description |
|--------|-------------|
| **Stressor values** | The first column contains raw stressor values (e.g., temperature in °C) |
| **Mean System Capacity** | Expected system capacity (0-100%) at each stressor level |
| **SD** | Standard deviation representing uncertainty in the relationship |
| **low.limit / up.limit** | Absolute bounds for the system capacity (used to constrain random sampling) |

### Visualizing the Dose-Response Curve

The figure below illustrates how to interpret a stressor-response curve. The optimal temperature for this species is around 16°C (100% capacity). As temperature increases beyond the optimum, capacity declines rapidly, reaching 0% at temperatures above 24°C.

```{r, dr_sample, echo = FALSE, message=FALSE, warning=FALSE, fig.width = 6, fig.height = 4, fig.cap="Example stressor-response curve showing how temperature affects system capacity. Error bars show standard deviation; grey shading shows the allowable range.", fig.align = 'center'}
  {
    plot(dft$Temperature, dft$msc, type ='o', pch = 19,
       xlab = "Temperature (°C)", ylab = "Mean System Capacity (%)",
       ylim = c(0, 110))

    x <- c(dft$Temperature, rev(dft$Temperature))
    y <- c(dft$up.limit, rev(dft$low.limit))

    polygon(x, y, col = "lightgrey", border = NA)

    points(dft$Temperature, dft$msc, type ='o', pch = 19)
    uppr <- ifelse(dft$msc + dft$SD > 100, 100, dft$msc + dft$SD)

    arrows(dft$Temperature, dft$msc - dft$SD, dft$Temperature, uppr, length=0.05, angle=90, code=3)
  }
```

**Key points about the uncertainty:**

- The **error bars** (standard deviation) represent natural variability or uncertainty in the relationship
- The **grey shading** shows the absolute bounds - system capacity values will never fall outside this range
- During Monte Carlo simulations, values are sampled from this distribution to propagate uncertainty through the analysis

## Loading the Stressor-Response Workbook

Use the `StressorResponseWorkbook()` function to import and validate your data:

```{r, import_dr, echo = TRUE, message = FALSE, warning = FALSE}
# Load the example stressor-response workbook included with CEMPRA
filename_sr <- system.file("extdata", "stressor_response_fixed_ARTR.xlsx", package = "CEMPRA")

# Import and validate the workbook
sr_wb_dat <- StressorResponseWorkbook(filename = filename_sr)

# The function returns a list with two components:
names(sr_wb_dat)
```

The returned object contains:

- `main_sheet`: The Main worksheet data as a data frame
- `sr_dat`: A list of data frames, one for each stressor's dose-response curve


# Stressor-Magnitude Workbook

While the stressor-response workbook defines *how* stressors affect the system, the stressor-magnitude workbook tells us *what the actual stressor levels are* at each location.

This workbook contains location-specific measurements or estimates of each stressor, along with uncertainty bounds.

## Workbook Structure

```{r, sm_demo1, echo = FALSE, message=FALSE, warning=FALSE}
  filename_rm <- system.file("extdata", "stressor_magnitude_unc_ARTR.xlsx", package = "CEMPRA")
  smw <- StressorMagnitudeWorkbook(filename = filename_rm, scenario_worksheet = "natural_unc")
  smw <- smw[which(smw$HUC_ID == 1702010107), ]
  smw$NAME <- 'Example Location'
  df <- data.frame(head(smw, 4))
  df %>%
  kbl(caption = "Example stressor-magnitude data showing stressor levels at one location") %>%
  kable_classic(full_width = F, html_font = "Cambria")
```

Each row represents one stressor at one location:

| Column | Description |
|--------|-------------|
| **HUC_ID** | Unique numeric identifier for the location (originally Hydrological Unit Code, but can be any ID system) |
| **NAME** | Human-readable location name (for display purposes) |
| **Stressor** | Stressor name (must exactly match the stressor-response workbook) |
| **Stressor_cat** | Stressor category (must match the stressor-response workbook) |
| **Mean** | Mean/expected value of the stressor at this location |
| **SD** | Standard deviation representing measurement uncertainty or natural variability |
| **Distribution** | Sampling distribution: `normal` or `lognormal` |
| **Low_Limit / Up_Limit** | Bounds for the stressor values (prevents unrealistic samples) |

## Loading the Stressor-Magnitude Workbook

Use the `StressorMagnitudeWorkbook()` function to import your location data:

```{r, import_sm, echo = TRUE, message = FALSE, warning = FALSE}
# Load the example stressor-magnitude workbook
filename_rm <- system.file("extdata", "stressor_magnitude_unc_ARTR.xlsx", package = "CEMPRA")

# Import the workbook - note that you specify which scenario worksheet to use
smw <- StressorMagnitudeWorkbook(filename = filename_rm, scenario_worksheet = "natural_unc")

# Preview the column names
names(smw)
```

**Tip:** The stressor-magnitude workbook can contain multiple scenario worksheets (e.g., "current", "restored", "future_climate"). This allows you to compare system capacity under different management scenarios without creating multiple files.


# Calculating System Capacity

Now that we understand the input data, let's calculate system capacity. We'll start with a single stressor to understand the mechanics, then run the full model with all stressors.

## Step 1: Generate Mean Response Curves

Before calculating system capacity, we need to process the stressor-response data into interpolated curves. The `mean_Response()` function creates smooth curves from the discrete data points in your worksheets:

```{r, csc_demo, echo = TRUE, message = FALSE, warning = FALSE}
# Create interpolated response curves for all stressors
mean.resp.list <- mean_Response(
  n.stressors = nrow(sr_wb_dat$main_sheet),
  str.list = sr_wb_dat$sr_dat,
  main = sr_wb_dat$main_sheet
)
```

This creates a list of response functions that can convert any stressor value to a system capacity score.

## Step 2: Calculate Capacity for a Single Stressor

Let's calculate system capacity for temperature at a hypothetical location. This helps illustrate what the model is doing before we run the full analysis.

```{r, single_stressor, echo = TRUE, message = FALSE, warning = FALSE}
# Extract data for the "Temperature_adult" stressor
f.main.df <- sr_wb_dat$main_sheet[which(sr_wb_dat$main_sheet$Stressors == "Temperature_adult"), ]
f.stressor.df <- sr_wb_dat$sr_dat[["Temperature_adult"]]
f.mean.resp.list <- mean.resp.list[[which(sr_wb_dat$main_sheet$Stressors == "Temperature_adult")]]

# Create sample data for a hypothetical location
# Mean temperature = 14°C, SD = 3°C
smw_sample <- data.frame(
  HUC_ID = 0,
  NAME = "Example Location",
  Stressor = "Temperature_adult",
  Stressor_cat = "Temperature",
  Mean = 14,
  SD = 3,
  Distribution = "normal",
  Low_Limit = 4,
  Up_Limit = 20
)

# Calculate system capacity with 100 Monte Carlo simulations
test_sc <- SystemCapacity(
  f.dose.df = smw_sample,
  f.main.df = f.main.df,
  f.stressor.df = f.stressor.df,
  f.mean.resp.list = f.mean.resp.list,
  n.sims = 100
)

# Convert to percentage
sys_cap <- round(test_sc$sys.cap * 100, 3)
```

## Understanding the Monte Carlo Results

The model runs multiple simulations (100 in this case) to capture uncertainty. In each simulation:

1. A temperature value is randomly drawn from the specified distribution (mean=14°C, SD=3°C)
2. That temperature is converted to a system capacity using the dose-response curve
3. Additional uncertainty from the dose-response relationship is incorporated

The result is a distribution of possible system capacity values:

```{r, tmsc_plot, echo = TRUE, message=FALSE, warning=FALSE, fig.width = 6, fig.height = 4, fig.cap="Distribution of system capacity values for temperature at the example location. The spread reflects uncertainty in both the temperature measurement and the dose-response relationship.", fig.align = 'center'}
hist(sys_cap,
     xlab = "System Capacity (%)",
     main = "System Capacity Distribution - Temperature Only",
     col = "steelblue",
     border = "white")
abline(v = mean(sys_cap), col = "red", lwd = 2, lty = 2)
legend("topright", legend = paste("Mean =", round(mean(sys_cap), 1), "%"),
       col = "red", lty = 2, lwd = 2, bty = "n")
```

At a mean temperature of 14°C, the system capacity is relatively high (around 80%), which matches our dose-response curve where 14°C corresponds to good habitat conditions.


# Running the Full Joe Model

Now let's run the complete Joe Model, which:

1. Calculates system capacity for **every stressor** at **every location**
2. Combines individual stressor effects into a **cumulative system capacity** score
3. Repeats this process across multiple Monte Carlo simulations to quantify uncertainty

## Loading the Data

```{r, jmd_load, echo = TRUE, message = FALSE, warning = FALSE}
# Load both workbooks
filename_rm <- system.file("extdata", "stressor_magnitude_unc_ARTR.xlsx", package = "CEMPRA")
filename_sr <- system.file("extdata", "stressor_response_fixed_ARTR.xlsx", package = "CEMPRA")

dose <- StressorMagnitudeWorkbook(filename = filename_rm, scenario_worksheet = "natural_unc")
sr_wb_dat <- StressorResponseWorkbook(filename = filename_sr)

# Check how many locations and stressors we have
cat("Number of locations:", length(unique(dose$HUC_ID)), "\n")
cat("Number of stressors:", nrow(sr_wb_dat$main_sheet), "\n")
```

## Running the Model

The `JoeModel_Run()` function handles all the calculations:

```{r, jmd, echo = TRUE, message = FALSE, warning = FALSE}
# Run the Joe Model
jmr <- JoeModel_Run(
  dose = dose,           # Stressor magnitude data
  sr_wb_dat = sr_wb_dat, # Stressor-response relationships
  MC_sims = 10           # Number of Monte Carlo simulations (use more for final analysis)
)

# Examine the output structure
names(jmr)
```

## Understanding the Output

The Joe Model returns a list with two key components:

| Component | Description |
|-----------|-------------|
| `ce.df` | **Cumulative Effects** - Overall system capacity for each location and simulation |
| `sc.dose.df` | **Stressor-specific Capacity** - System capacity broken down by individual stressor |

Let's visualize the cumulative system capacity across all locations and simulations:

```{r, msc_plot, echo = TRUE, message=FALSE, warning=FALSE, fig.width = 6, fig.height = 4, fig.cap="Distribution of cumulative system capacity across all locations and Monte Carlo simulations. Lower values indicate locations where multiple stressors are limiting habitat quality.", fig.align = 'center'}
# Extract cumulative system capacity (as percentage)
msc <- jmr$ce.df$CE * 100

hist(msc,
     xlab = "Cumulative System Capacity (%)",
     main = "Joe Model Results - All Locations",
     col = "steelblue",
     border = "white",
     breaks = 20)
abline(v = mean(msc), col = "red", lwd = 2, lty = 2)
legend("topleft",
       legend = c(paste("Mean =", round(mean(msc), 1), "%"),
                  paste("Min =", round(min(msc), 1), "%"),
                  paste("Max =", round(max(msc), 1), "%")),
       bty = "n")
```

## Interpreting the Results

The histogram shows the distribution of cumulative system capacity values. Key insights:

- **High values (>70%)**: Locations with good habitat conditions where few stressors are limiting
- **Medium values (40-70%)**: Locations where some stressors are degrading habitat quality
- **Low values (<40%)**: Heavily impacted locations where multiple stressors are limiting capacity

The spread in values reflects both:

1. **Spatial variation**: Different locations have different stressor levels
2. **Uncertainty**: Monte Carlo sampling propagates uncertainty from both stressor measurements and dose-response relationships


# Next Steps

This tutorial covered the basics of running the Joe Model for a single scenario. To learn more about:

- **Running multiple scenarios**: See [Tutorial 2: Joe Model Batch Run](a02-joe-model-batch-run.html)
- **Linking to population dynamics**: See [Tutorial 3: Population Model Overview](a03-population-model.html)
- **Sensitivity analysis**: See [Tutorial 6: Population Model Evaluation](a06-population-model-sensitivity.html)

For detailed guidance on preparing your own data, see the [CEMPRA Documentation](https://mattjbayly.github.io/CEMPRA_documentation/).
