---
title: "6. Population Model Evaluation"
output: rmarkdown::html_vignette
date: '`r Sys.Date()`'
vignette: >
  %\VignetteIndexEntry{6. Population Model Evaluation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
# browseVignettes()
```

## Overview

We can compare the sensitivity of parameters within the population model. This is useful to understand parameters that are more or less sensitive to perturbations. In this example we will swap out different life cycle profiles for populations of Chinook Salmon and see how the sensitivity of input parameters changes our results. We will use stressor-response and stressor-magnitude values from for Nanaimo River Chinook salmon, but we will swap out the life cycle profile data for the following populations:

* Nanaimo River Summer Chinook (vital rates from DFO-RAMs).
* Chehalis River Spring/Fall Chinook in coastal Washington (using parameters published from Beechie et al., 2021).
* Columbia River Chinook from Wenatchee River (using parameters from Honea et al., 2009: Freshwater Biology).
* Nicola River spring-run Chinook from the interior of BC (DFO-RAMs 2021).


## Load Input Data

In this example we will load the stressor-response, stressor magnitude file and life cycles file. The polygon geometry is not needed to run the population model. Reset `filename_` to point to a local file on your computer. We will load the Nanaimo River Chinook salmon data as a starting point.


```{r m0}
  library(CEMPRA)

  # Stressor Magnitude
  fname <- system.file("extdata/nanaimo/stressor_magnitude_nanaimo.xlsx", package = "CEMPRA")
  dose <- StressorMagnitudeWorkbook(filename = fname)
  dose$SD <- 0

  # Stressor Response
  fname <- system.file("extdata/nanaimo/stressor_response_nanaimo.xlsx", package = "CEMPRA")
  sr_wb_dat <- StressorResponseWorkbook(filename = fname)

  # Load location and habitat capacities
  filename <- system.file("extdata/nanaimo/habitat_capacities_nanaimo.csv", package = "CEMPRA")
  habitat_capcities <- read.csv(filename, stringsAsFactors = FALSE)
    
  # Load the life cycle parameters
  filename_lc <- system.file("extdata/nanaimo/species_profiles/nanaimo_comp_ocean_summer.csv", package = "CEMPRA")
  life_cycle_params <- read.csv(filename_lc)
    
    
```

##  Target Populations

The following habitat capacities files lists the target populations. The density-dependent constraints esimtate capacities in the absence of any stressors and assume that 100% of the stream area is suitable for spawning (e.g., gravel coverage = 100% and spawning suitability = 100%). We have specific stressors targetting capacities that will shink these numbers down to more realistic estimates.

1. Lower Nanaimo River Fall-Run Chinook
2. Lower Nanaimo River Summer-Run Chinook
3. Upper Nanaimo River Spring/Summer-Run Chinook


```{r m1}
print(habitat_capcities)

# Focus analysis on the Lower Nanaimo River Summer-Run Chinook (1)
HUC_ID <- 2

```

## Run the Population Model

Next, we can run the population model for a number of years `n_years` and batch replicates `MC_sims`. 

```{r m2}
    
    data <- PopulationModel_Run(
        dose = dose,
        sr_wb_dat = sr_wb_dat,
        life_cycle_params = life_cycle_params,
        HUC_ID = 2, # Lower Summer-run Chinook
        n_years = 200, # run for 200 years
        MC_sims = 5, # batch replicates
        habitat_dd_k = habitat_capcities # Population-specific carrying capacities
    )

    # Model returns object with all data nested in a list object
    names(data)
    # ce = with cumulative effect stressors
    
    # 5 batch replicates
    length(data$ce)
    
    # Look at replicate 2 of 5
    names(data$ce[[2]])
    
    # We want to review the N object that contains the population abundance in each stage class
    tail(round(data$ce[[2]]$N, 0))

```

## Visualize the Time Series Projections
Simple plots can be included to visualize trends

```{r m3, out.width='100%', fig.align='center', dpi=300, fig.height=5}

spawners <- pop_model_get_spawners(data = data,
                                   life_stage = "spawners",
                                   life_cycle_params = life_cycle_params)

head(spawners)

library(dplyr)
summary_data <- spawners %>%
  filter(year > 50) %>%
  group_by(location_id, year) %>%
  summarize(
    mean_spawners = mean(spawners),
    sd_spawners = sd(spawners),
    .groups = 'drop'
  )
head(summary_data)

library(ggplot2)
# Create the ggplot
ggplot(summary_data, aes(x = year, y = mean_spawners, group = location_id)) +
  geom_line(color = "blue", size = 1) +  # Line for mean spawners
  geom_ribbon(
    aes(ymin = mean_spawners - sd_spawners, ymax = mean_spawners + sd_spawners),
    fill = "blue",
    alpha = 0.2
  ) +  # Shading for ±1 standard deviation
  labs(title = "Time Series of Spawners", x = "Simulation Year", y = "Number of Spawners (N)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5), text = element_text(size = 12))

```


# Location and life stage-specific BH capacities
Run an Example with BH carrying capacities



